\documentclass[UserManual.tex]{subfiles}
\begin{document}
\setcounter{section}{4}
\section{Tuning the Emulator}\label{sec:emulator}

\subsection{Summary}

Smooth emulator finds a sample set of Taylor expansion coefficients that reproduce a set of observables at a set of training points. The process of finding those coefficients is referred to as ``tuning''. For a given observable, a particular sample set of coefficients gives the following emulated function:
\begin{align*}\eqnumber
E(\vec{\theta})&=\sum_{\vec{n}, s.t. \sum_in_i\le {\rm MaxRank}}d(\vec{n})A_{\vec{n}}
\left(\frac{\theta_1}{\Lambda}\right)^{n_1}
\left(\frac{\theta_2}{\Lambda}\right)^{n_2}
\cdots.
\end{align*}
Here, $\theta_1\theta_2\cdots$ represent the original model parameters, $\vec{X}$, but are scaled. If their initial prior is uniform, they are scaled so that their priors range from -1 to +1, and if they have  Gaussian priors, they are scaled so that their variance is one third. The degeneracy factor, $d(\vec{n})$ is the number of different ways to sum the powers $n_i$ to a given rank,
\begin{align*}\eqnumber
d(\vec{n})=\sqrt{\frac{(n_1+n_2+\cdots)!}{n_1!n_2!\cdots}}.
\end{align*}
As described in Sec. \ref{sec:theory}, the coefficients are chosen weighted by the distribution,
\begin{eqnarray}
P(\vec{A})=\prod_n\frac{1}{\sqrt{2\pi\sigma_A^2}}e^{-A_n^2/2\sigma_A^2},
\end{eqnarray}
where $\sigma_A$ is varied to maximize the overall probability given the constraint of reproducing the training points. More discussion is provided in Sec. \ref{sec:theory}. Whereas {\it Smooth Emulator} does a nice job of finding an optimum value for $\sigma_A$, the smoothness parameter $\Lambda$ is unfortunately difficult to optimize. For the moment, this is treated purely as prior knowledge, or expectation. If the User expects the full model to be very smooth, i.e. the quadratic contributions to be much smaller than the linear contributions and so on, a larger value (e.g 3.0), might be chosen. If the full-model output might be almost wavy, then a smaller value (e.g. 1.5) might be chosen. The emulator uncertainties will be smaller for larger $\Lambda$.

By setting parameters, as described below, {\it Smooth Emulator} can tuned one of three different ways
\begin{itemize}\itemsep=0pt
\item[a)] Find the optimum set of coefficients. If evaluated a the training points, the emulator will exactly produce the full model. when it predicts the observable at a new $\vec{\theta}$ it provides an uncertainty.
\item[b)] If a Monte Carlo tuning method is chosen, the emulator finds a predetermined number of sets of coefficients, where each set of coefficients provides a function that exactly reproduces the real model at the training points. The User sets the number of sets of coefficients, typically of order $N_{\rm sample}\approx 10$, in the parameter file. Away from the training points, the uncertainty of the emulator is represented by the spread of the values amongst the $N_{\rm sample}$ predictions.
\item[c)] The third mode also provides $N_{\rm sample}$ predictions, but rather than exactly reproducing the training values the emulator merely comes close to the training points with a distribution $\sim e^{-\Delta y^2/2\sigma_y^2}$, where $\sigma_y$ represents the random error of the full model. This mode should be chosen if the full model has significant random error, and especially if the training points are close to one another.
\end{itemize}
Method (a) is by far the quickest, and will probably be used the most often. 

If methods (b) or (c) are chosen {\it Smooth Emulator} solves for the $N_{\rm sample}$ sets of coefficients from the training data, then stores $N_{\rm sample}$ sets of coefficients, along with the averaged coefficients in files for later use. If (a) is chosen, {\it Smooth Emulator} stores the set of ``best'' coefficients along with some other arrays used for rapid calculation of the uncertainty. {\it Smooth Emulator} can emulate either the full-model observables directly, or their principal components. Training the emulator follows the same steps for either approach. 

The executables based on {\it Smooth Emulator} are located in the User's {\tt \$\{MY\_LOCAL\}/bin} directory. Examples of such executables are {\tt smoothy\_tune} or {\tt smoothy\_calcobs}. These functions must be executed from within the User's project directory. 

In the following subsections, we first review the format for each of the required input files, then describe how to run {\it Smooth Emulator}, how its output is stored, and how to switch PCA observables for real observables.

\subsection{Preparing Files for {\it Smooth Emulator}}

Before training the emulator, one must first run the full model at a given set of training points. In addition to a parameter file (described in the next sub-section), which sets numerous options, the User must provide the following:
\begin{enumerate}\itemsep=0pt
\item A file listing the names of observables and an estimate of the variance of each observable throughout the model-parameter space, $\sigma_A$. This file is named {\tt Info/observable\_info.txt}, where the path is relative to the project directory. The file might look like
{\tt
\begin{verbatim}
 obsname1  12.3  
 obsname2  23.4 
 obsname3  34.5
 obsname4  45.6  
\end{verbatim}}
\vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
The initial $\sigma_A$ is only relevant if one is using one of the Monte Carlo tuning methods, (b) or (c) above, as it provides an initial guess for the parameter $\sigma_A$ above.

\item A file listing the names of the model parameters that also describes their priors. This file is {\tt Info/modelpar\_info.txt}. The file might have the following form:
{\tt
\begin{verbatim}
parname1 uniform   0      1.0E-3
parname2 uniform   -50.0  100.0
parname3 gaussian  0      24.6
parname4 uniform   30.0   50.0
\end{verbatim}}
\vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
If the prior is {\tt uniform} the two following numbers provide the minimum and maximum of the interval. If the prior is {\tt gaussian} the two subsequent values represent the center and r.m.s. width of the Gaussian. This same file was required for running {\it Simplex Sampler}.

\item A list of the model-parameter values, $\vec{\theta}_{\rm train}$, at each training point. These points can be generated by {\it Simplex Sampler}, as described in Sec. \ref{sec:simplex}, or they can be generated by hand. If the number of full-model runs performed is $N_{\rm train}$, Smooth emulator requires files for each run. Each file is named {\tt \$\{MODEL\_RUN\_DIRNAME\}/runI/mod\_parameters.txt}, where $0\le I<N_{\rm train}$, and $I$ denotes the point in  parameter space for the $I^{\rm th}$ full-model training run. The directory {\tt \$\{MODEL\_RUN\_DIRNAME\}/} is typically {\tt \$\{MY\_PROJECT\}/modelruns}, but can be defined otherwise (see below). For example the file {\tt modelruns/run0/mod\_parameters.txt} might look like
{\tt
\begin{verbatim}
 parname1  8.34E-4
 parname2  -30.5375
 parname3  36.238
 parname4  39.34
\end{verbatim}}
\vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\

\item At each training point, the User must provide the full model's value for each observable. In the same directory where the model-parameter values are stored, {\it Smooth Emulator} requires the observables calculated at the training points mentioned above. This information is provided in {\tt \$\{MODEL\_RUN\_DIRNAME\}/runI/obs.txt}. An example of such a file is:
{\tt
\begin{verbatim}
 obsname1  -51.4645   2.5
 obsname2  166.837    0.9
 obsname3  -47.9877   0.0
 obsname4  -2.34526   0.03
\end{verbatim}}
\vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
 \vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
The first number is the calculated value of the observable, and the second is the random error. This is only the random error, i.e. that which represents that if the model were rerun at the same training point, the value might be different. This should only be non-zero if the full-model has some Monte Carlo feature. For example, the full model might involve simulating a small number of events. Other types of uncertainty are accounted for by including them into the experimental uncertainty.

Once the emulator is tuned and before it is applied to a Markov Chain investigation of the likelihood, the software needs know the experimental measurement and uncertainty. That information must be entered in the {\tt Info/experimental\_info.txt} file. The file should have the format:
{\tt
\begin{verbatim}
 obsname1  -12.93    0.95   0.5
 obsname2  159.3     3.0    2.4
 obsname3  -61.2.    1.52   0.9
 obsname4  -1.875    0.075  0.03
\end{verbatim}}
\vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
 \vspace*{-16pt}
 \hspace*{28pt}$\vdots$\\
The first number is the measured value and the second is the experimentally reported uncertainty. The third number is the uncertainty inherent to the theory, due to missing physics. For example, even if a model has all the parameters set to the exact value, e.g. some parameter of the standard value, the full-model can't be expected to exactly reproduce a correct measurement given that some physics is likely missing from the full model. For the MCMC software, the relevant uncertainty incorporates both, and only the combination of both, added in quadrature, affects the outcome. We emphasize that this last file is not needed to train and tune the emulator. It is needed once one performs the MCMC search of parameter space. 


\end{enumerate}




\subsection{{\it Smooth Emulator} Parameters (not model parameters!)}

{\it Smooth Emulator} requires a parameter file. This can be located anywhere, as it will be specified on the command line when running {\it Smooth Emulator}, but is typically {\tt parameters/emulator\_parameters.txt}. The parameter file is simply a list, of parameter names followed by values. 

{\tt
\begin{verbatim}
 #SmoothEmulator_LogFileName smoothlog.txt # comment out for interactive running
  SmoothEmulator_LAMBDA 2.0 #  Smoothness parameter
  SmoothEmulator_MAXRANK 5
  SmoothEmulator_ConstrainA0 true
  SmoothEmulator_ModelRunDirName modelruns
  SmoothEmulator_TrainingPts 0-27
  SmoothEmulator_UsePCA   false
  SmoothEmulator_TuneChooseExact true
 #
 # These are only used if you are using MCMC tuning rather than Exact method
  SmoothEmulator_TuneChooseMCMC false # set false if NPars<5
  SmoothEmulator_MCMC_NASample 8  # No. of coefficient samples
  SmoothEmulator_MCMC_StepSize 0.01
  SmoothEmulator_TuneChooseMCMCPerfect false #
  SmoothEmulator_MCMC_UseSigmaY false # Emulator only fits training data to within model random error
  SmoothEmulator_MCMC_CutoffA false # Used only if SigmaA constrained by SigmaA0
  SmoothEmulator_MCMC_SigmaAStepSize 1.0  #
  SmoothEmulator_MCMC_NMC 20000  # Steps between samples 
 #
 # This is for the MCMC search of parameter space (not for the emulator tuning)
 MCMC_METROPOLIS_STEPSIZE 0.01
 RANDY_SEED.  1234
\end{verbatim}
}
If any of these parameters are missing from the parameters file, {\it Smooth Emulator} will assign a default value.

\begin{enumerate}\itemsep 0pt
\item {\bf SmoothEmulator\_LogFileName}\\
If this is commented out, as it is in the example above, {\it Smooth Emulator}'s main output will be directed to the screen and the program will run interactively. Otherwise, the output will be recorded in the specified file. Most often, one will wish the program to run interactively.

\item {\bf SmoothEmulator\_LAMBDA}\\
This is the smoothness parameter $\Lambda$. It sets the relative importance of terms of various rank. If $\Lambda$ is unity or less, it suggests that the Taylor expansion converges slowly. The default is 3.

\item {\bf SmoothEmulator\_MAXRANK}\\
As {\it Smooth Emulator} assumes a Taylor expansion, this the maximum power of $\theta^n$ that is considered. Higher values require more coefficients, which in turn, slows down the tuning process. The default is 4.

\item {\bf SmoothEmulator\_ConstrainA0}\\
The coefficients in the Taylor expansion are assumed to have some weight,
\[
W(A_i)=\frac{1}{\sqrt{2\pi\sigma_A^2}}e^{-A_i^2/2\sigma_A^2}.
\]
The term $\sigma_A$ is allowed to vary during the tuning to maximize the likelihood of the expansion. If the User wishes to exempt the lowest term, i.e. the constant term in the Taylor expansion from the weight, the User may set {\tt SmoothEmulator\_ConstrainA0} to {\tt false}. The default is {\tt false}.

\item {\bf SmoothEmulator\_ModelRunDirName}\\
This gives the directory in which the training data from the full model runs is stored. The default is {\tt modelruns}, which is the same default {\tt Simplex Sampler} uses for writing the coordinates of the training points.

\item {\bf SmoothEmulator\_TrainingPts}\\
This lists which full-model training runs SmoothEmulator will use to train the emulator. This provides the User with the flexibility to use some subset for training, as may be the case when testing the accuracy. The default is ``1''. An example the User might enter could be\\
{\tt ~SmoothEmulator\_TrainingPts  0-4,13,15}\\
This would choose the training information from the directories {\tt run0,run1,run2,run3,run4,run13} and {\tt run15}, which would be found in the directory denoted by the {\tt SmoothEmulator\_ModelRunDirName} parameter.

\item {\bf SmoothEmulator\_UsePCA}\\
By default, this is set to {\tt false}. If one wishes to emulate the PCA observables, i.e. those that are linear combinations of the real observables, this should be set to true. One must then be sure to have run the pca decomposition programs first. For more, see Sec. \ref{sec:pca}. 

\item {\bf SmoothEmulator\_TuneExact}\\
This is set to {\tt true} by default. In this case {\it Smooth Emulator} finds the optimum set of coefficients and also records some other arrays necessary for calculating the emulator uncertainty. This is tuning type (a) above.

\item {\bf RANDY\_SEED}\\
This sets the seed for the random number generator. If the line is commented out, it will be set to {\tt std::time(NULL)}.

\#\#\# None of the parameters below are relevant if {\tt SmoothEmulator\_TuneExact} is {\tt true}.

\item {\bf SmoothEmulator\_TuneChooseMCMC}\\
This is tuning type (b) above. Rather than finding the optimum set of coefficients, {\it Smooth Emulator} finds $N_{\rm sample}$ sets of coefficients consistent with the probability. All the sets exactly reproduce training observables. As this runs as a Markov chain, the independence of the sample may require a large number of steps between samplings. The default is {\tt false}.

\item {\bf SmoothEmulator\_MCMC\_NASample}\\
{\it Smooth Emulator} finds $N_{\rm sample}$ sets of coefficients. Each set reproduces the training points, but differs away from the training points. Setting $N_{\rm sample}\sim 10$ should reasonably represent the uncertainty of the emulator. The default is set at 8.

\item {\bf SmoothEmulator\_MCMC\_NMC}\\
This is the number of Markov Chain steps between samples. A larger number is required if the samplings are to be independent. The default is 20,000.

\item {\bf SmoothEmulator\_MCMC\_StepSize}\\
This is the size of the MCMC stepsize in $\theta$ space. MCMC approaches are most efficient if the success probability is near 0.5. If the probability is much higher, then the step size should be increased, and if it is much lower, the step size should be decreased. This affects only the efficiency, not the accuracy. The default of 0.01.

\item {\bf SmoothEmulator\_TuneChooseMCMCPerfect}\\
If there are a small number of model parameters, perhaps less than a half dozen, then rather than performing a Markov Chain one can choose the coefficients by a keep-or-reject method. The advantage of this approach is that $N_{\rm sample}$ coefficients are perfectly independent. The disadvantage is that the the percentage of ``keeps'' falls rapidly with an increasing number of parameters. For larger numbers of parameters it is usually more efficient to set this to its default, {\tt false}. 

\item {\bf SmoothEmulator\_MCMC\_UseSigmaY}\\
If the real model has significant random noise, the emulator should not be constrained to exactly reproduce the observables at the training points. This is tuning type (c) above. 

\item {\bf SmoothEmulator\_MCMC\_CutoffA}\\
This applies an additional multiplicative weight to the weight for $A$ above:
\[
W(A_i)_{\rm additional}=\frac{1}{1+\frac{1}{4}\frac{A_i^2}{\sigma_{A}^2}}.
\] 
Here $\sigma_{A0}$ is the initial guess for the spread. This can safeguard against the width $\sigma_A$ drifting off to arbitrarily large values. Unless necessary, it is recommended to leave this at the default, {\tt false}.

\end{enumerate}

\subsection{Running the {\it Smooth Emulator} Program}

The source code for the {\it Smooth Emulator} main program can be found in the {\tt \$\{MY\_LOCAL\}/main\_programs/} directory. This directory contains source code for several main programs. They are separated from the bulk of the software, which is in the {\tt GITHOME\_BAND/SmoothEmulator/software/} directory. The main programs are designed so that the User can easily copy and edit them to create versions that might be more appropriate to the User's specific needs. When compiled, from the {\tt \$\{MY\_LOCAL\}/build/} directory, the executables appear in the {\tt \$\{MY\_LOCAL\}/bin/} directory. Two of the source codes that come with the distributions are \$\{MY\_LOCAL\} and {\tt \$\{MY\_LOCAL\}/main\_programs/smoothy\_calc\_main.cc}. Once compiled the corresponding executables are {\tt \$\{MY\_LOCAL\}/bin/smoothy\_tune} and {\tt \$\{MY\_LOCAL\}/bin/smoothy\_calcobs}. A 
{\tt
\begin{verbatim}
using namespace std;
int main(int argc,char *argv[]){
    if(argc!=2){
        printf("Usage smoothy emulator parameter filename");
        exit(1);
    }
    CparameterMap *parmap=new CparameterMap();
    parmap->ReadParsFromFile(string(argv[1]));
    CSmoothMaster master(parmap);
    master.ReadTrainingInfo();
    master.GenerateCoefficientSamples();
    master.WriteCoefficientsAllY();
    return 0;
}
\end{verbatim}
}
Similarly, there is a code {\tt \$\{MY\_LOCAL\}/main\_programs/smoothy\_calcobs\_main.cc}, which provides an example of how one might read the coefficients and generate predictions for the emulator at specfied points in parameter space.

From within the {\tt \$\{MY\_LOCAL\}/build/} directory, one can compile the two programs with the commands:
{\tt
\begin{verbatim}
 MY_LOCAL/build % cmake .
 MY_LOCAL/build % make smoothy_tune
 MY_LOCAL/build % make smoothy_calcobs
\end{verbatim}
}
The executables {\tt smoothy\_tune} and {\tt smoothy\_calcobs} should now appear in the {\tt \$\{MY\_LOCAL\}/bin/} directory. Assuming the {\tt bin/} directory has been added to the User's path, the User may switch to the User's project directory, and enter the command
{\tt
\begin{verbatim}
  ~/MY_PROJECT % smoothy_tune PARAMETERS/MY_PARAMETERS.TXT
\end{verbatim}
}
Here {\tt PARAMETERS/MY\_PARAMETERS.TXT} can be replaced by the User, but is typically {\tt parameters/emulator\_parameters.txt}.

The program will write the Taylor coefficients for the $N_{\rm sample}$ samples to files in the {\tt coefficients} directory.  The coefficients for each observable are given in separate subdirectories, named by the observables, i.e. {\tt coefficients/OBS\_NAME/sampleI.txt}. Here, , where {\tt OBS\_NAME} is the name for each observable, and if there are $N_{\rm sample}$ sets of coefficients, $0\le I<N_{\rm sample}$. Along with the coefficients, in the same directory {\it Smooth Emulator} writes a file for each observable. These files are named {\tt coefficients/OBS\_NAME/meta.txt}.  This file provides information, such as the maximum rank and net number of model parameters, to make it possible to read the coefficients later on.

{\it Smooth Emulator} will output lines describing its progress, either to the screen or to a file, as specified by the {\tt SmoothEmulator\_LogFile} parameter described above. This output includes a report on the percentage of steps in the MCMC program that were successful. The line {\tt BestLogP/Ndof} describes the weight used to evaluate the likelihood of a coefficients sample. This value should roughly plateau once the Metropolis procedure has settled on the most likely region.

For later us, e.g. when performing the MCMC to sampler the posterior, the User would need to generate predictions for specified values of the parameters. The executable {\tt \$\{MY\_LOCAL\}/bin/smoothy\_calcobs}, is such an example. It is compiled from the main program, {\tt \$\{MY\_LOCAL\}/main\_programs/smoothy\_calcobs.cc}:
{\tt
\begin{verbatim}
int main(int argc,char *argv[]){
    if(argc!=2){
        CLog::Info("Usage smoothy_calcobs emulator parameter filename");
        exit(1);
    }
    CparameterMap *parmap=new CparameterMap();
    parmap->ReadParsFromFile(string(argv[1]));
    CSmoothMaster master(parmap);
    // Reads Emulator Coefficients for all observables
    master.ReadCoefficientsAllY();
    master.priorinfo->PrintInfo();
    //modpars carries info about single point
    CModelParameters *modpars=new CModelParameters(master.priorinfo);
    // Prompt user for model parameter values
    vector<double> X(modpars->NModelPars);
    for(int ipar=0;ipar<modpars->NModelPars;ipar++){
        cout << "Enter value for " << master.priorinfo->GetName(ipar) << ":\n";
        cin >> X[ipar];
    }
    modpars->SetX(X);
    //  Calc Observables Y[iy] for X
    CObservableInfo *obsinfo=new CObservableInfo("Info/Observable_Info.txt");
    vector<double> Y(obsinfo->NObservables);
    vector<double> SigmaY(obsinfo->NObservables);
    master.CalcAllY(modpars,Y,SigmaY);
    for(int iY=0;iY<obsinfo->NObservables;iY++){
        cout << obsinfo->GetName(iY) << " = " << Y[iY] << " +/- " << SigmaY[iY] << endl;
    }
    return 0;

\end{verbatim}
}
The User can hopefully use this template programs as a base for calling {\it Smooth Emulator} to calculate the emulator values for a specified point in space. Note that {\tt SigmaY} is the emulator uncertainty, not that from experiment or from the theoretical model.


\end{document}
